---
title: "Test"
author: "Ewy Mathe"
date: "8/14/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Simulating a dataset
```{r}
set.seed(123)
age <- rnorm(2000, mean = 60, sd = 10)
length(age)
range(age)
sd(age)
mean(age)
```

Now, let's visualize results with a boxplot and a histogram.

```{r}
par(mfrow=c(2,2))
hist(age)
hist(age, breaks=100)
boxplot(age, main = "Boxplot of Age",
        xlab = "Individuals", ylab = "Age")
```

Now let's create our variable for gender, with 800 "F" and 1200 "M"

```{r}
gender <- c(rep("F",800),rep("M",1200))
set.seed(234)
finalgender <- sample(gender, 2000, replace=F)
table(finalgender)
length(finalgender)
head(finalgender)
```

Now let us work on creating a variable called location: 500 "LocA", 1,200 "LocB", 800 "LocC"

```{r}
set.seed(345)
location <- c(rep("LocA",500),rep("LocB",1200), rep("LocC",300))
table(location)
finallocation <- sample(location, length(location), replace=FALSE) 
table(finallocation)
head(finallocation)
```

Lastly, create the variable "caco", that has 1,000 "Cases" and 1,000 "Controls"
```{r}
set.seed(456)
caco <- c(rep("Cases", 1000), rep("Controls", 1000))
table(caco)

finalcaco <- sample(caco, length(caco), replace=FALSE) 
table(finalcaco)
head(finalcaco)
```

Now creating a data frame for my simultated data:
```{r}
mydata <- data.frame(age=age, caco=finalcaco,
                     gender=finalgender, location=finallocation)
summary(mydata)
```

Select a subset of 1,000 individuals, 500 cases and 500 controls, that have similar data distributions between the cases and controls.

```{r}
set.seed(5)
myrows <- sample(1:nrow(mydata), 1000, replace=FALSE)

subsetmydata <- mydata[myrows,]
hist(subsetmydata[,"age"],breaks=100)

cases <- subsetmydata[which(subsetmydata$caco == "Cases"),]
controls <- subsetmydata[which(subsetmydata$caco == "Controls"),]
```

Let's check that the distributions are the same in cases and controls:
```{r}
summary(cases)
summary(controls)
```

Be a good R citizen:
```{r}
sessionInfo()
```
